Asymmetric relationships and their effects on coexistence

### 2023-02-03 paper/Paper.Rnw
finished Methods and accompanying SI sections, at least for now

%***DAN: Next steps as of 2023 02 01:
%DONE Fill in the sup mat section for the theory for the beta fecundities lottery model
%fill in the efficient computation section for that same model
%DONE write this section, here, in the main text

@@

%Then on to the diatom stuff.

%G3
To achieve goal G3 from the Introduction, i.e., to explore, using the lottery model, the likelihood that species can coexist with
ATAs but not without, or vice versa, we plotted GWR, $\olr_{i \bs i}=\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$, 
as well as
GWR minus its ATA contribution, $\olr_{i \bs i} -\Delta_i^{[EC]}$, again for species $i=1$, the weaker competitor. 
For fixed values of $\sigma$ and $\delta$ and for left-tail associated noise, 
we were interested in the values of $\mu_1-\mu_2$ for which one of these
quantities is positive and the other is negative; this set of values characterizes the species which can coexist
with ATAs, but not without, or vice versa. We define the ``trait-difference space'' of values $\mu_1-\mu_2$. 
Points of the trait-difference space characterize possible combinations of traits of the two 
competing species in our model. For fixed values of 
$\delta$ and $\sigma$, we defined the subset, $\psi_{\text{ATA}}(\delta,\sigma)$, of 
trait-difference space to be the
values $\mu_1 - \mu_2$ for which GWR, $\olr_{i \bs i}$, is positive. This 
is the subspace for which the two lottery model species coexist in the presence of ATAs.
We also defined the subset, $\psi_{\lnot\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the values
$\mu_1 - \mu_2$ for which the metric $\olr_{i \bs i} -\Delta_i^{[EC]}$ is positive.
This is the subspace for which the two lottery model species coexist in the absence of 
ATAs. The symbol $\lnot$ denotes logical negation, so $\lnot \text{ATA}$ indicates the
absence of ATAs. We defined the ``ATA rescue space" associated with given values of $\delta$ and $\sigma$ to be the 
set difference $\psi_{\text{ATA}}(\delta,\sigma) \setminus \psi_{\lnot\text{ATA}}(\delta,\sigma)$, i.e., the
values $\mu_1 - \mu_2$ for which species coexist in the presence of ATAs, but not in their absence. 
We define ``ATA exclusion space'' to be the set difference 
$\psi_{\lnot\text{ATA}}(\delta,\sigma) \setminus \psi_{\text{ATA}}(\delta,\sigma)$,
i.e., $\mu_1 - \mu_2$ values for which coexistence occurs in the absence, but not the presence, of ATAs.
On plots of $\olr_{i \bs i}$ and $\olr_{i \bs i} -\Delta_i^{[EC]}$
against $\mu_1 - \mu_2$, we highlighted the ATA rescue and exclusion spaces. We plotted against
$\mu_1 - \mu_2$ for all combinations of $\delta=0.2, 0.4, 0.6, 0.8$ and $\sigma=2,4,5,6$.

%G4
To achieve goal G4 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we use a laboratory chemostat
system of freshwater diatoms explored by \cite{Descamps_2005}.
Those authors used measurements of growth rates of \emph{Fragilaria crotonensis} and 
\emph{Cyclotella pseudostelligera} over a range of temperatures to parametrize a resource
competition model. The model is 
\begin{align}
\frac{dS}{dt} &= D(S_0 - S) - Q_1x_1 \frac{V_1 S}{K_1 + S} - Q_2 x_2 \frac{V_2 S}{K_2 + S} \\
\frac{dx_j}{dt} &= x_j \frac{V_j S}{K_j+S}-D x_j, 
\end{align}
for $j=1,2$, where $S$ is a extracellular silicate concentration in the chemostat, $x_1$ is the population density
of \emph{F. crotonensis} and $x_2$ is that of \emph{C. pseudostelligera}, 
$D$ is the chemostat outflow rate, $S_0$ is concentration of silicates in the chemostat
inflow, the $Q_j$ are amounts of silicate per cell, the $V_j$ are maximum population growth rates, and
the $K_j$ are half-saturation constants for the dependence of growth rates on nutrient concentration.
The temperature-dependent parameters $Q_j$, $V_j$, and $K_j$ were obtained by 
\cite{Descamps_2005} through batch experiments and curve fitting or interpolation.
\cite{Descamps_2005} showed that permanent coexistence of the two species occurs when 
temperature fluctuates periodically, 
$\theta(t)=\theta_0 +  a\sin(2\pi t/P)$, 
with parameters $\theta_0 = 18^\circ$C, $a = 6^\circ$C, $P = 60$d. 
\cite{Ellner_2016, Ellner_2019} broke down coexistence into contributing mechanisms
according to (\ref{eq:previous_decomp}) for these same values of
$\theta_0$, $a$ and $P$. We further decompose
coexistence according to (\ref{eq:full_decomp}), and we do so for ranges of values
of $\theta_0$, $a$ and $P$. In contrast to the lottery model case, for which simplifying assumptions
made it sufficient to decompose only the GWR of species 1, for this
example we considered $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$ for both $i=1$, $j=2$ and for
$i=2$, $j=1$. Further details of the model setup are in SI section \ref{SI-sect:diatom_model}
and specifics of how our decomposition extends to this empirical case are in 
SI section \ref{SI-sect:Empircal_app}.

To display results, we plotted the terms of (\ref{eq:full_decomp}) against one of the parameters 
$\theta_0$, $a$ and $P$ at a time, while keeping the other two fixed at the values
used by \cite{Descamps_2005}. We also generated two-dimensional plots, where
two of the parameters $\theta_0$, $a$ and $P$ were varied and the third was kept fixed
at the value used by \cite{Descamps_2005}. For the two-dimensional plots, the quantity  
$\Delta_i^{[EC]}/\text{GWR}$ was displayed using color, to convey the 
importance of the ATA contribution relative to the whole GWR. Larger magnitudes
of this quantity indicate greater importance of ATAs. The case
$\Delta_i^{[EC]}>\text{GWR}>0$ (which produces $\Delta_i^{[EC]}/\text{GWR}>1$)
corresponds to ATA rescue, and the case $\Delta_i^{[EC]}<\text{GWR}<0$
(which also produces $\Delta_i^{[EC]}/\text{GWR}>1$) corresponds to ATA exclusion.
Thus the condition $\Delta_i^{[EC]}/\text{GWR}>1$ encompasses the cases for which ATA
exclusion or rescue occurred, though the condition does not delineate, by itself,
which of these was the case.

>>>>

%Trait different space stuff. I don't think I need it anymore but keeping around until I finish writing the Discussion,
%in case some of it turns out to be useful there.
% To achieve goal G3 from the Introduction, i.e., to explore, using the lottery model, the likelihood that species can coexist with
% ATAs but not without, or vice versa, we plotted GWR, $\olr_{i \bs i}=\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$, 
% as well as
% GWR minus its ATA contribution, $\olr_{i \bs i} -\Delta_i^{[EC]}$, again for species $i=1$, the weaker competitor. 
% For fixed values of $\sigma$ and $\delta$ and for left-tail associated noise, 
% we were interested in the values of $\mu_1-\mu_2$ for which one of these
% quantities is positive and the other is negative; this set of values characterizes the species which can coexist
% with ATAs, but not without, or vice versa. We define the ``trait-difference space'' of values $\mu_1-\mu_2$. 
% Points of the trait-difference space characterize possible combinations of traits of the two 
% competing species in our model. For fixed values of 
% $\delta$ and $\sigma$, we defined the subset, $\psi_{\text{ATA}}(\delta,\sigma)$, of 
% trait-difference space to be the
% values $\mu_1 - \mu_2$ for which GWR, $\olr_{i \bs i}$, is positive. This 
% is the subspace for which the two lottery model species coexist in the presence of ATAs.
% We also defined the subset, $\psi_{\lnot\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the values
% $\mu_1 - \mu_2$ for which the metric $\olr_{i \bs i} -\Delta_i^{[EC]}$ is positive.
% This is the subspace for which the two lottery model species coexist in the absence of 
% ATAs. The symbol $\lnot$ denotes logical negation, so $\lnot \text{ATA}$ indicates the
% absence of ATAs. We defined the ``ATA rescue space" associated with given values of $\delta$ and $\sigma$ to be the 
% set difference $\psi_{\text{ATA}}(\delta,\sigma) \setminus \psi_{\lnot\text{ATA}}(\delta,\sigma)$, i.e., the
% values $\mu_1 - \mu_2$ for which species coexist in the presence of ATAs, but not in their absence. 
% We define ``ATA exclusion space'' to be the set difference 
% $\psi_{\lnot\text{ATA}}(\delta,\sigma) \setminus \psi_{\text{ATA}}(\delta,\sigma)$,
% i.e., $\mu_1 - \mu_2$ values for which coexistence occurs in the absence, but not the presence, of ATAs.
% On plots of $\olr_{i \bs i}$ and $\olr_{i \bs i} -\Delta_i^{[EC]}$
% against $\mu_1 - \mu_2$, we highlighted the ATA rescue and exclusion spaces. We plotted against
% $\mu_1 - \mu_2$ for all combinations of $\delta=0.2, 0.4, 0.6, 0.8$ and $\sigma=2,4,5,6$.

@@

To achieve goal G3 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we used a laboratory chemostat
system of freshwater diatoms explored by \cite{Descamps_2005}.
Those authors used measurements of growth rates of \emph{Fragilaria crotonensis} and 
\emph{Cyclotella pseudostelligera} over a range of temperatures to parametrize a resource
competition model. The model is 
\begin{align}
\frac{dS}{dt} &= D(S_0 - S) - Q_1x_1 \frac{V_1 S}{K_1 + S} - Q_2 x_2 \frac{V_2 S}{K_2 + S} \\
\frac{dx_j}{dt} &= x_j \frac{V_j S}{K_j+S}-D x_j, 
\end{align}
for $j=1,2$, where $S$ is a extracellular silicate concentration in the chemostat, $x_1$ is the population density
of \emph{F. crotonensis} and $x_2$ is that of \emph{C. pseudostelligera}, 
$D$ is the chemostat outflow rate, $S_0$ is concentration of silicates in the chemostat
inflow, the $Q_j$ are amounts of silicate per cell, the $V_j$ are maximum population growth rates, and
the $K_j$ are half-saturation constants for the dependence of growth rates on nutrient concentration.
The temperature-dependent parameters $Q_j$, $V_j$, and $K_j$ were obtained by 
\cite{Descamps_2005} through batch experiments and curve fitting or interpolation.
\cite{Descamps_2005} showed that permanent coexistence of the two species occurs when 
temperature fluctuates periodically, 
$\theta(t)=\theta_0 +  a\sin(2\pi t/P)$, 
with parameters $\theta_0 = 18^\circ$C, $a = 6^\circ$C, $P = 60$d. 
\cite{Ellner_2016, Ellner_2019} broke down coexistence into contributing mechanisms
according to (\ref{eq:previous_decomp}) for these same values of
$\theta_0$, $a$ and $P$. We further decompose
coexistence according to (\ref{eq:full_decomp}), and we do so for ranges of values
of $\theta_0$, $a$ and $P$. In contrast to the lottery model case, for which simplifying assumptions
made it sufficient to decompose only the GWR of species 1, for this
example we considered $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$ for both $i=1$, $j=2$ and for
$i=2$, $j=1$. Further details of the model setup are in SI section \ref{SI-sect:diatom_model}
and specifics of how our decomposition extends to this empirical case are in 
SI section \ref{SI-sect:Empircal_app}.

\subsection{How results were plotted for goal G3}

%***DAN: This para will likely have to be revised depending on how Jasmin revises the diatom figure presentation
To display results for goal G3, we plotted the terms of (\ref{eq:full_decomp}) against one of the parameters 
$\theta_0$, $a$ and $P$ at a time, while keeping the other two fixed at the values
used by \cite{Descamps_2005}. We also generated two-dimensional plots, where
two of the parameters $\theta_0$, $a$ and $P$ were varied and the third was kept fixed
at the value used by \cite{Descamps_2005}. For the two-dimensional plots, the quantity  
$\Delta_i^{[EC]}/\text{GWR}$ was displayed using color, to convey the 
importance of the ATA contribution relative to the whole GWR. Larger magnitudes
of this quantity indicate greater importance of ATAs. The case
$\Delta_i^{[EC]}>\text{GWR}>0$ (which produces $\Delta_i^{[EC]}/\text{GWR}>1$)
corresponds to ATA rescue, and the case $\Delta_i^{[EC]}<\text{GWR}<0$
(which also produces $\Delta_i^{[EC]}/\text{GWR}>1$) corresponds to ATA exclusion.
Thus the condition $\Delta_i^{[EC]}/\text{GWR}>1$ encompasses the cases for which ATA
exclusion or rescue occurred, though the condition does not delineate, by itself,
which of these was the case.
